---
title: "Beautiful Tables"
---


We are going to reproduce some of the tables of this paper _Seasonal Changes in Plankton Food Web Structure and Carbon Dioxide Flux from Southern California Reservoirs_ <https://doi.org/10.1371/journal.pone.0140464>


```{r label='install_load'}
# install.packages("gt")
library(gt)
library(tidyverse)
```

## Load the data

```{r label='read_in_data'}
data <- read.csv("data/Adamczyk_PLOSdata.csv")
```


## Have a look at the data

```{r label='summary'}
data |> glimpse()
```


Let's have a look at the first rows of our dataset

```{r}
data |> head() |> gt()
```

Now let's create a summary table of the average air temperature for each basin

```{r}
data_air <- temp_basin <- data |>
  select(year, lake, air_temp_c) |>
  group_by(year, lake) |>
    summarise(average_air_temp = mean(air_temp_c))

data_air
```

## Now let's make this a nice table

```{r}
gt_air <-
  data_air |>
  gt() |>
  tab_header(
    title = "Average air temperature (C) per lake and year",
    subtitle = "major reservoirs in San Diego County"
  )

# Show the gt Table
gt_air
```

## Add reference

```{r}
gt_air <-
  data_air |>
  gt() |>
  tab_header(
    title = "Average air temperature (C) per lake and year",
    subtitle = "major reservoirs in San Diego County"
  ) |>
 tab_source_note(
    source_note = md("Source: _Adamczyk EM, Shurin JB (2015) Seasonal Changes in Plankton Food Web Structure and Carbon Dioxide Flux from Southern California Reservoirs. PLoS ONE 10(10): e0140464. https://doi.org/10.1371/journal.pone.0140464_")
  )

# Show the gt Table
gt_air
```


## With more measurements


```{r}
data_measurements <- temp_basin <- data |>
  select(year, lake, air_temp_c, pco2_atmosphere_ppm, h2o_temp_c, ph, pco2_water_ppm) |>
  group_by(year, lake) |>
    summarise(avg_air_temp = mean(air_temp_c),
              avg_pco2_atmosphere_ppm = mean(pco2_atmosphere_ppm, na.rm=TRUE),
              avg_h2o_temp_c = mean(h2o_temp_c, na.rm=TRUE),
              avg_ph = mean(ph, na.rm=TRUE),
              avg_pco2_water_ppm = mean(pco2_water_ppm, na.rm=TRUE)
    )

```


# Add title and Footnote


```{r}
gt_measurements <-
  data_measurements |>
  gt() |>
  tab_header(
    title = "Average measurements per lake and year",
    subtitle = "major reservoirs in San Diego County"
  ) |>
 tab_source_note(
    source_note = md("Source: _Adamczyk EM, Shurin JB (2015) Seasonal Changes in Plankton Food Web Structure and Carbon Dioxide Flux from Southern California Reservoirs. PLoS ONE 10(10): e0140464. https://doi.org/10.1371/journal.pone.0140464_")
  )

# Show the gt Table
gt_measurements
```

## Adding headers for multiple columns

```{r}
gt_measurements <-
  data_measurements |>
  gt() |>
  tab_header(
    title = "Average measurements per lake and year",
    subtitle = "major reservoirs in San Diego County"
  ) |>
 tab_source_note(
    source_note = md("Source: _Adamczyk EM, Shurin JB (2015) Seasonal Changes in Plankton Food Web Structure and Carbon Dioxide Flux from Southern California Reservoirs. PLoS ONE 10(10): e0140464. https://doi.org/10.1371/journal.pone.0140464_")
  ) |>
  tab_spanner(
    label = "Air",
    columns = c(avg_air_temp, avg_pco2_atmosphere_ppm)
  ) |>
  tab_spanner(
    label = "Water",
    columns = c(avg_h2o_temp_c, avg_ph, avg_pco2_water_ppm)
  )


# Show the gt Table
gt_measurements
```

## More styling on sub headers

```{r}
gt_measurements <-
  data_measurements |>
  gt() |>
  tab_header(
    title = "Average measurements per lake and year",
    subtitle = "major reservoirs in San Diego County"
  ) |>
 tab_source_note(
    source_note = md("Source: _Adamczyk EM, Shurin JB (2015) Seasonal Changes in Plankton Food Web Structure and Carbon Dioxide Flux from Southern California Reservoirs. PLoS ONE 10(10): e0140464. https://doi.org/10.1371/journal.pone.0140464_")
  ) |>
  tab_spanner(
    label = md("**Air**"),
    columns = c(avg_air_temp, avg_pco2_atmosphere_ppm)
  ) |>
  tab_spanner(
    label = md("**Water**"),
    columns = c(avg_h2o_temp_c, avg_ph, avg_pco2_water_ppm)
  )


# Show the gt Table
gt_measurements
```

## Simplify the column names

```{r}
gt_measurements  <- gt_measurements |> cols_label(
    avg_air_temp = "Temp (C)",
    avg_pco2_atmosphere_ppm = "pC02 (ppm)",
    avg_h2o_temp_c = "Temp (C)",
    avg_ph = 'pH',
    avg_pco2_water_ppm = "pC02 (ppm)"
  )

gt_measurements
```

## Limiting to 2 decimals

```{r}
gt_measurements |>  fmt_number(
    columns = everything(),
    decimals = 2,
    use_seps = FALSE
  )
```

## Styling


```{r}
gt_measurements |> opt_stylize(style = 6, color = 'gray')
```

