---
title: "Beautiful Tables with `gt`"
---

## A few words on the `gt` package


With `gt` you can construct a wide variety of useful tables with a cohesive set of table parts. Here is the main workflow to transform a data frame into a beautiful table:

![https://gt.rstudio.com/reference/figures/gt_workflow_diagram.svg](https://gt.rstudio.com/reference/figures/gt_workflow_diagram.svg)

An `gt` object as various elements. These include the table **header**, the **stub**, the **column labels** and **spanner column labels**, the **table body**, and the **table footer**.

![https://gt.rstudio.com/reference/figures/gt_parts_of_a_table.svg](https://gt.rstudio.com/reference/figures/gt_parts_of_a_table.svg)

Checkout the `gt` website for more information: <https://gt.rstudio.com/articles/gt.html>

# Our example 

We are going to reproduce some of the tables of this paper _Seasonal Changes in Plankton Food Web Structure and Carbon Dioxide Flux from Southern California Reservoirs_ <https://doi.org/10.1371/journal.pone.0140464> and create new ones using the data archived on Dryad <https://doi.org/10.5061/dryad.6tn4h>.


## Load the libraries

```{r label='install_load'}
#| message: FALSE
 
# install.packages("gt")
library(gt)
library(tidyverse)
```

## Load the data

```{r label='read_in_data'}
data <- read.csv("data/Adamczyk_PLOSdata.csv")
```


## Have a look at the data

```{r label='summary'}
data |> glimpse()
```


Let's have a look at the first rows of our dataset, but this time we are going to use the `gt` package

```{r}
data |> 
  head() |>
  gt()
```

## A simple table

Now let's create a summary table of the average air temperature for each basin

```{r}
#| message: FALSE

data_air <- data |>
  select(year, lake, air_temp_c) |>
  group_by(year, lake) |>
    summarise(average_air_temp = mean(air_temp_c))

data_air
```



## Now let's make this a nice table

```{r}
gt_air <-
  data_air |>
  gt() 

# Show the gt Table
gt_air
```

## Add a Title
```{r}
#| message: FALSE

gt_air |> 
  tab_header(
    title = "Average measurements per lake and year",
    subtitle = "major reservoirs in San Diego County"
  ) 
```

## Working with a more complex table

```{r}
#| message: FALSE

data_measurements <- data |>
  select(year, lake, air_temp_c, pco2_atmosphere_ppm, h2o_temp_c, ph, pco2_water_ppm) |>
  group_by(year, lake) |>
    summarise(avg_air_temp = mean(air_temp_c),
              avg_pco2_atmosphere_ppm = mean(pco2_atmosphere_ppm, na.rm=TRUE),
              avg_h2o_temp_c = mean(h2o_temp_c, na.rm=TRUE),
              avg_ph = mean(ph, na.rm=TRUE),
              avg_pco2_water_ppm = mean(pco2_water_ppm, na.rm=TRUE)
    )

data_measurements |>  gt()
```

## Add better column names

```{r}
gt_measurements <- data_measurements |>
  gt() |>
  cols_label(
    avg_air_temp = "Air Temp (C)",
    avg_pco2_atmosphere_ppm = "Air pC02 (ppm)",
    avg_h2o_temp_c = "Water Temp (C)",
    avg_ph = 'Water pH',
    avg_pco2_water_ppm = "Water pC02 (ppm)"
  )


# Show the gt Table
gt_measurements
```


## Adding headers for multiple columns

```{r}
gt_measurements <- gt_measurements |>
  tab_spanner(
    label = md("**Air**"),
    columns = c(avg_air_temp, avg_pco2_atmosphere_ppm)
  ) |>
  tab_spanner(
    label = md("**Water**"),
    columns = c(avg_h2o_temp_c, avg_ph, avg_pco2_water_ppm)
  ) |>
  cols_label(
    avg_air_temp = "Temp (C)",
    avg_pco2_atmosphere_ppm = "pC02 (ppm)",
    avg_h2o_temp_c = "Temp (C)",
    avg_ph = 'pH',
    avg_pco2_water_ppm = "pC02 (ppm)"
  )

# Show the gt Table
gt_measurements
```



## Limiting values to 2 decimals

```{r}
gt_measurements <- gt_measurements |>
    fmt_number(
    columns = everything(),
    decimals = 2,
    use_seps = FALSE
  )


# Show the gt Table
gt_measurements
```

## Put it all together

### Add title and Footnote


```{r}
gt_measurements <- gt_measurements |>
  tab_header(
    title = "Average measurements per lake and year",
    subtitle = "major reservoirs in San Diego County"
  ) |>
 tab_source_note(
    source_note = md("Source: _Adamczyk EM, Shurin JB (2015) Seasonal Changes in Plankton Food Web Structure and Carbon Dioxide Flux from Southern California Reservoirs. PLoS ONE 10(10): e0140464. <https://doi.org/10.1371/journal.pone.0140464>_")
  )

# Show the gt Table
gt_measurements
```


## Styling


```{r}
gt_measurements |> opt_stylize(style = 6, color = 'gray')
```


There are 36 combinations of style and color to choose from: <https://gt.rstudio.com/reference/opt_stylize.html?q=opt_stylize#examples>

And of course you can do conditional cell formatting: <https://gt.rstudio.com/reference/tab_style.html>
