---
title: "Beautiful Tables for Models"
---

## Introduction to `gtsummary`

Next, we'll dive into the `gtsummary` package, an extension of `gt` that can incorporate model outputs into tables.

As with all R packages, we'll first need to install the package, then call it for use during this session.

```{r label='install_load'}
# install.packages("gtsummary")
library(gtsummary)
```



## Sample data

We'll use the same sample data from [Adamczyk & Shurin 2016](https://datadryad.org/dataset/doi:10.5061/dryad.6tn4h) as during the `gt` section.

If you haven't already, download [Adamczyk_PLOSdata.csv](https://datadryad.org/downloads/file_stream/33666) to the `data` folder in this repository and then import the csv, assigning to a dataframe called `data`:

```{r label='read_in_data'}
data <- read.csv("data/Adamczyk_PLOSdata.csv")
```

## Descriptive statistics

The base R approach to getting summary statistics for a dataset results in a thorough, but not well-formatted for publication, table.

```{r label='base_summary'}
summary(data)
```
The equivalent summary statistics function from `gtsummary` is `tbl_summary()`.

```{r label='tbl_summary'}
tbl_summary(data)
```

The default summary statistic presented depends on variable type. Categorical variables get counts and percentages while continuous variables get `median(first quartile, third quartile)`. Any NA values are listed in a separate row as "unknown".

These defaults are customizable. For instance, if we wanted only a subset of variables and for continuous variables to include minimum, median, and maximum:

```{r label='customize_descriptive'}
data %>%  
  tbl_summary(
    include = c(lake, h2o_temp_c, chlorophyll_a_ugL),
    statistic = list(
      #change just way continuous vars are presented
      all_continuous() ~ "{median} ({min}, {max})"
    ),
  )
```

_Note: `gtsummary` uses |> as a pipe operator. The `tidyverse` %>% will also work and is used here._

We can also produce crosstab tables, for instance with descriptive statistics for selected variables for each lake.

```{r label='crosstab'}
data %>% 
  tbl_summary(
    by = lake,
    include = c(h2o_temp_c, chlorophyll_a_ugL),
    statistic = list(
      #change just way continuous vars are presented
      all_continuous() ~ "{median} ({min}, {max})"
    )
  )
```

And customize the table with better labels and updated NA indicator:

```{r label='customize_labels'}
data %>% 
  tbl_summary(
    by = lake,
    include = c(h2o_temp_c, chlorophyll_a_ugL),
    statistic = list(
      #change just way continuous vars are presented
      all_continuous() ~ "{median} ({min}, {max})"
    ),
    label = list(h2o_temp_c = "Water Temperature (C)", chlorophyll_a_ugL = "Chlorophyll a (ugL)"),
    missing_text = "Missing values"
  )
```

## Statistical tests

If we wanted to run statistical analyses on this data, a usual approach would be to use a separate R package to run the test and then extract values of interest, like coefficients and p-values, to put into a table. `gtsummary` streamlines this process with an all-in-one approach to formatting statistical test outputs into a table.

There are many standard [statistical tests integrated into gtsummary](https://www.danieldsjoberg.com/gtsummary/reference/tests.html) including t-test, ANOVA, chi-square, regression, survey sample methods, and more.

Let's look at a couple examples in practice.

First, let's say we wanted to run a one-way ANOVA to compare whether chlorophyll is significantly different between the three lakes in 2013.

```{r}
# data %>% 
#   filter(year == 2013) %>% 
# ggplot(aes(x = julian_date, y = chlorophyll_a_ugL))+
#   geom_point(aes(color = lake))
```
```{r}
oneway.test(chlorophyll_a_ugL ~ as.factor(lake), data = data)
```
If we wanted to report these results, we'd have some work to do to paste this into a table format. Or, we can do the same analysis using the [`tbl_summary`](https://www.danieldsjoberg.com/gtsummary/articles/tbl_summary.html) function from `gtsummary` to do this for us!

> The tbl_summary() function calculates descriptive statistics for continuous, categorical, and dichotomous variables in R, and presents the results in a beautiful, customizable summary table ready for publication (for example, Table 1 or demographic tables).

Let's run the same one-way ANOVA as above:

```{r}
data %>% 
  #specify what variables we want in our table: here, we want a row for chlorophyll with columns for lakes
  tbl_summary(by = lake, include = c(chlorophyll_a_ugL)) |>
  #specify what test to perform on which variables, to get p-values from
  #here, a one-way ANOVA on chlorphyll
  add_p(test = chlorophyll_a_ugL ~ "oneway.test")
```

#### [STILL DRAFTING]

Thinking through variables and order matters.

If we don't have table with by=lakes, the test doesn't know which groups to compare for anova

```{r}
#| error: true
data %>% 
  #no specification that we are grouping our data into lake-specific columns
  tbl_summary(include = c(chlorophyll_a_ugL)) |>
  add_p(test = chlorophyll_a_ugL ~ "oneway.test")
```

Also, if we don't specify a test, `add_p()` will use the [default tests](The default test used in `add_p()` primarily depends on these factors:), which depend on whether the data is continuous or categorical, how many categories, etc.

If we don't specify anova, the analysis will default to a kruskal-wallis rank sum test.

>The default test used in add_p() primarily depends on these factors:
> * whether the variable is categorical/dichotomous vs continuous
> * number of levels in the tbl_summary(by) variable
> * whether the add_p(group) argument is specified
> * whether the add_p(adj.vars) argument is specified

```{r}
data %>% 
  tbl_summary(by= lake, include = c(chlorophyll_a_ugL)) |>
  add_p()
```


## Additional resources

* [Primary website for gtsummary](https://www.danieldsjoberg.com/gtsummary/) including how to [cite use of the package](https://www.danieldsjoberg.com/gtsummary/authors.html#citation).
* gtsummary [reference manual](https://cran.r-project.org/web/packages/gtsummary/refman/gtsummary.html) with complete details on functions and arguments.
* Examples of [gtsummary in practice from R Graph Gallery](https://r-graph-gallery.com/package/gtsummary.html). 

## Data reference

Adamczyk, Emily M.; Shurin, Jonathan B. (2016). Data from: Seasonal changes in plankton food web structure and carbon dioxide flux from Southern California reservoirs [Dataset]. Dryad. [https://doi.org/10.5061/dryad.6tn4h](https://doi.org/10.5061/dryad.6tn4h)